---
title: "Course Project - 07 Regression"
author: "Joe Nguyen"
date: "19 November 2015"
output: html_document
---

```{r, echo=FALSE}
# Change working directory
dirBase <- "/home/joe/Documents/01-coursera/01-data-science"
dirWorking <- "/07-regression"
setwd(file.path(dirBase, dirWorking))
rm(list = ls())
```


# Executive Summary

This report aims to explore two questions:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions

Cars are more fuel efficient if their miles per gallon (MPG) rating is lower.

## Explore `mtcars` dataset

```{r}
data("mtcars")
head(mtcars,2)
tail(mtcars,2)
```

Variable `am` is transmission type (0 = automatic, 1 = manual) (from R documentation `?mtcars`).

```{r, fig.align='center', fig.height=4, fig.width=4}
require(ggplot2)
mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual"))
ggplot(aes(am, mpg), data = mtcars) +
    geom_violin() + 
    ggtitle("MPG Distribution for Transmission Type") + 
    xlab("Transmission Type") + ylab("MPG")
```


## Model Fitting

We can treat `am` as a factor variable with two levels and perform regression with the linear model:

$$ Y_i = \beta_0 + X_{i1} \beta_1 + \epsilon_i $$

Using least squares, $\hat{\beta}_0$ is the mean estimate for "automatic" and $\hat{\beta}_0 + \hat{\beta}_1$ is the mean estimate for "manual".


```{r}
modLm <- lm(mpg ~ am - 1, data = mtcars)
summary(modLm)

```


